Radio Crickets: Chirping Jets from Black Hole Binaries Entering their Gravitational Wave Inspiral

### 2015-12-02 Paper/paper.tex
Correct typos as pointed out by referee

The brighter part of the jet (intensity within factor of 10 from the
maximum) in Figure~\ref{fig:jet_mdot1.0} (right panel) has separations
of $\sim 1$ mas.  It should therefore be possible to resolve the jet
at an angular resolution of $\sim 2$ mas.  (In the next section, we
discuss an algorithm to fit the jet model to observational data.)  If
it is aligned closer to the line of sight, the forward jet is
magnified due to apparant superluminal motion.  This will make it
possible to resolve jets of objects that are farther away than the
example in Figure~\ref{fig:jet_mdot1.0}.  Binaries with higher masses
have large orbital periods that will result is a helix with wider
seperation between its cycles.  These objects would therefore be
easier to detect.  We have not modelled the absolute surface
brightness of jets, only relative variation in the jet brightness (as
shown in Figure~\ref{fig:jet_mdot1.0}), but due to the $(1+z)^{-4}$
dimming, it will be exceedingly difficult to detect such jets at
higher redshifts.

@@

From observations of chirping jets, we can infer the dynamical
parameters of the binary BH by first fitting a purely geometric model
to the observed jet morphology.  To create a model of this kind, we
generalise the simple helical model considered by
\citet{2014MNRAS.445.1370K} and \citet{2001NewA....6...61N} to write
\begin{align}
  z &= x^{-\alpha} \frac{B}{2\pi} u \cos{u}, \\
  y &= x^{-\alpha} \frac{B}{2\pi} u \sin{u}, \\
  x &= \frac{A}{2\pi} u \left(\frac{u}{2\pi}\right)^\beta.
\end{align}
Here $u$ is the azimuthal angle around the jet axis, $x$ and $y$ axes
are in the plane of the sky, and the $z$ axis points to the observer.
The $x^{-\alpha}$ term in the first two equations captures the
widening opening angle of the jet, while the $(u/2\pi)^{\beta}$ term
in the last equation parameterises the chirp.  This model can be fit
to an observed jet on the sky by using two rotation matrices,
$\mathcal{R}_\lambda$ and $\mathcal{R}_i$, which rotate the model
in the above equtions to match with the jet precession axis and the
jet symmetry axis.  These matrices are given by

\begin{flalign}
\mathrm{R}_{i} \!=\!
\begin{pmatrix}
\cos i & 0 & \sin i \\ 
0 & 1 & 0 \\ 
-\sin i & 0 & \cos i
\end{pmatrix}
\label{riota}
\end{flalign}
and
\begin{flalign}
\mathcal{R}_{\lambda}\!=\!
\begin{pmatrix}
\cos \lambda{_0} & -\sin \lambda{_0} & 0 \\ 
\sin \lambda{_0} & \cos \lambda{_0} & 0 \\ 
0 & 0 & 1%
\end{pmatrix}.
\label{rlambda}
\end{flalign}
Further, we can parameterise the ``twist'' in the jet by requiring
that $i$ and $\lambda$ vary with $x$.  Without loss of generality,
we take this variation in $i$ and parameterise it as
\begin{equation}
  i = \gamma x + i_0.
\end{equation}
We thus arrive at a model with 7 parameters ($\alpha$, $\beta$,
$\gamma$, $A$, $B$, $i_0$, $\lambda_0$), which can be fit to any
observed jet, such as that in Figure~\ref{fig:jet_mdot1.0}.  Note that
setting $\alpha$ and $\beta$ equal to $0$ yields the usual helical jet
morphology, without the features discussed in this paper.  Thus, a
non-zero value of these two parameters distinguishes chirpy jets from
non-chirpy ones.

>>>>

The brighter part of the jet (intensity within factor of 10 from the
maximum) in Figure~\ref{fig:jet_mdot1.0} (right panel) has separations
of $\sim 1$ mas.  It should therefore be possible to resolve the jet
at an angular resolution of $\sim 2$ mas.  (In the next section, we
discuss an algorithm to fit the jet model to observational data.)  If
it is aligned closer to the line of sight, the forward jet is
magnified due to apparant superluminal motion.  This will make it
possible to resolve jets of objects that are farther away than the
example in Figure~\ref{fig:jet_mdot1.0}.  Binaries with higher masses
have large orbital periods that will result in a helix with wider
seperation between its cycles.  These objects would therefore be
easier to detect.  We have not modelled the absolute surface
brightness of jets, only relative variation in the jet brightness (as
shown in Figure~\ref{fig:jet_mdot1.0}), but due to the $(1+z)^{-4}$
dimming, it will be exceedingly difficult to detect such jets at
higher redshifts.

@@

From observations of chirping jets, we can infer the dynamical
parameters of the binary BH by first fitting a purely geometric model
to the observed jet morphology.  To create a model of this kind, we
generalise the simple helical model considered by
\citet{2014MNRAS.445.1370K} and \citet{2001NewA....6...61N} to write
\begin{align}
  z &= x^{-\alpha} \frac{B}{2\pi} u \cos{u}, \\
  y &= x^{-\alpha} \frac{B}{2\pi} u \sin{u}, \\
  x &= \frac{A}{2\pi} u \left(\frac{u}{2\pi}\right)^\beta.
\end{align}
Here $u$ is the azimuthal angle around the jet axis, $x$ and $y$ axes
are in the plane of the sky, and the $z$ axis points to the observer.
The $x^{-\alpha}$ term in the first two equations captures the
widening opening angle of the jet, while the $(u/2\pi)^{\beta}$ term
in the last equation parameterises the chirp.  This model can be
fitted to an observed jet on the sky by using two rotation matrices,
$\mathcal{R}_\lambda$ and $\mathcal{R}_i$, which rotate the model in
the above equtions to match with the jet precession axis and the jet
symmetry axis.  These matrices are given by

\begin{flalign}
\mathrm{R}_{i} \!=\!
\begin{pmatrix}
\cos i & 0 & \sin i \\ 
0 & 1 & 0 \\ 
-\sin i & 0 & \cos i
\end{pmatrix}
\label{riota}
\end{flalign}
and
\begin{flalign}
\mathcal{R}_{\lambda}\!=\!
\begin{pmatrix}
\cos \lambda{_0} & -\sin \lambda{_0} & 0 \\ 
\sin \lambda{_0} & \cos \lambda{_0} & 0 \\ 
0 & 0 & 1%
\end{pmatrix}.
\label{rlambda}
\end{flalign}
Further, we can parameterise the ``twist'' in the jet by requiring
that $i$ and $\lambda$ vary with $x$.  Without loss of generality,
we take this variation in $i$ and parameterise it as
\begin{equation}
  i = \gamma x + i_0.
\end{equation}
We thus arrive at a model with 7 parameters ($\alpha$, $\beta$,
$\gamma$, $A$, $B$, $i_0$, $\lambda_0$), which can be fitted to any
observed jet, such as that in Figure~\ref{fig:jet_mdot1.0}.  Note that
setting $\alpha$ and $\beta$ equal to $0$ yields the usual helical jet
morphology, without the features discussed in this paper.  Thus, a
non-zero value of these two parameters distinguishes chirpy jets from
non-chirpy ones.
